---
title: "Reproducible Research - Course Project 2"
author: "echf"
date: "11/8/2021"
output: html_document
---

## Exploring the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database - Health and Economic Impacts

#### Introduction

##### Storms and other severe weather events can cause both public health and economic problems for communities and 
##### municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to 
##### the extent possible is a key concern.


##### This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.  
##### This database tracks characteristics of major storms and weather events in the United States, including when and where they 
##### occur, as well as estimates of any fatalities, injuries, and property damage.

#### Data
##### The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce 
##### its size. You can download the file from the course web site: Storm Data 47Mb

 https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 

##### There is also some documentation of the database available. Here you will find how some of the variables are 
##### constructed/defined.


##### National Weather Service  Storm Data Documentation
    https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
##### National Climatic Data Center Storm Events FAQ
    https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf 
##### The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are 
##### generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more 
##### complete.






### Population health: 
 *  Number of injuries.
 *  Number of deaths.

### Economic impact: 
 *  Total cost of property damage.
 *  Total cost of crop damage.

### Data Processing
#### Load Libraries and prepare the R environment


```{r,echo = TRUE}
options(scipen = 1)  # Turn off scientific notations for numbers
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)
library(dplyr)
```



### Download the data


```{r,echo = TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "StormData.csv.bz2")
library(R.utils)
bunzip2("StormData.csv.bz2", "StormData.csv", overwrite=T, remove=F)
df_storm <- read.csv("StormData.csv")
names(df_storm)
str(df_storm)
```




#### Economic variables:
  
######  PROPDMG: approx. property damags
######  PROPDMGEXP: the units for property damage value
######  CROPDMG: approx. crop damages
######  CROPDMGEXP: the units for crop damage value
 
  
####   Events - target variable:
######     EVTYPE: weather event (Tornados, Wind, Snow, Flood, etc..)
######     Extract variables of interest from original data set:

####  Health variables:
  * INJURIES: approx. number of injuries
  * FATALITIES: approx. number of deaths
  * propdamage: property damage
  * cropdamage: Crop damage

```{r,echo = TRUE}
head(df_storm,10)
```


#### Extracting variables of interest for analysis of weather impact on health and economy
#### From a list of variables in storm data, these are columns of interest:
  


```{r,echo = TRUE}
df_storm2 <- df_storm %>% 
  select(c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")) %>%
  rename_all(tolower)
str(df_storm2)
```


##### Property damage estimates were entered as actual dollar amounts 

```{r,echo = TRUE}
sort(table(df_storm2$propdmgexp), decreasing = TRUE)[1:10]

sort(table(df_storm2$cropdmgexp), decreasing = TRUE)[1:10]

```
####  There is some mess in units, so we transform those variables in one unit (dollar) variable by the following rule: 
####  * K or k: thousand dollars (10^3) 
####  * M or m: million dollars (10^6) 
####  * B or b: billion dollars (10^9) 
####  * the rest would be consider as dollars 

####  Transform EVTYPE to all caps 
####  This will improve consistency across categories 

```{r,echo = TRUE}
df_storm2$evtype <- as.character(df_storm2$evtype)
df_storm2$evtype <- toupper(df_storm2$evtype)
```

####  Calculate property damage value 
####  Need to recode the exponent variable so we can multiply it by the propdmg value. 

```{r,echo = TRUE}  
df_storm2$propdmgexp <- toupper(df_storm2$propdmgexp)
df_storm2 <- mutate(df_storm2, propdmgexp2 = recode(propdmgexp,
                                             "1" = 10,
                                             "2" = 100,
                                             "3" = 1000,
                                             "4" = 10000,
                                             "5" = 100000,
                                             "6" = 1000000,
                                             "7" = 10000000,
                                             "8" = 100000000,
                                             "H" = 100,
                                             "K" = 1000,
                                             "M" = 1000000,
                                             "B" = 1000000000,
                                             .default = 1)) 
df_storm2 <- mutate(df_storm2, propdmgvalue = propdmg*propdmgexp2)
```


###  Calculate values

####  Calculate crop damage value  
####  Need to recode the exponent variable so we can multiply it by the cropdmg value. 
  
```{r,echo = TRUE}  
df_storm2$cropdmgexp <- toupper(df_storm2$cropdmgexp)
df_storm2 <- mutate(df_storm2, cropdmgexp2 = recode(cropdmgexp,  
                                                  "1" = 10,
                                                  "2" = 100,
                                                  "3" = 1000,
                                                  "4" = 10000,
                                                  "5" = 100000,
                                                  "6" = 1000000,
                                                  "7" = 10000000,
                                                  "8" = 100000000,
                                                  "H" = 100,
                                                  "K" = 1000,
                                                  "M" = 1000000,
                                                  "B" = 1000000000,
                                                 .default = 1))
df_storm2 <- mutate(df_storm2, cropdmgvalue = cropdmg*cropdmgexp2)
head(df_storm2,10)
```

####  Subset data for analyses 
####  Injury data 
  
```{r,echo = TRUE}  
injuries <- aggregate(injuries~evtype, data = df_storm2, FUN = sum) 
injuries <- injuries[order(injuries$injuries, decreasing = TRUE), ] 
injuries <- injuries[1:10, ] 
injuries
```

####  Fatality data 
```{r,echo = TRUE}  
fatalities <- aggregate(fatalities~evtype, data = df_storm2, FUN = sum) #aggregate fatalities by event type
fatalities <- fatalities[order(fatalities$fatalities, decreasing = TRUE), ] #order number of fatalities in decending order
fatalities <- fatalities[1:10, ] 
fatalities
```

####  Property damage data 
```{r,echo = TRUE}
propdamage <- aggregate(propdmgvalue~evtype, data = df_storm2, FUN = sum)
propdamage <- propdamage[order(propdamage$propdmgvalue, decreasing = TRUE), ]
propdamage <- propdamage[1:10, ]
propdamage
```

####  Crop damage data 
```{r,echo = TRUE}  
cropdamage <- aggregate(cropdmgvalue~evtype, data = df_storm2, FUN = sum)
cropdamage <- cropdamage[order(cropdamage$cropdmgvalue, decreasing = TRUE), ]
cropdamage <- cropdamage[1:10, ]
cropdamage
```


## The events that cause the most damage in relation to meteorological events in the United States of America are shown below. 


```{r,echo = TRUE} 

injuriesPlot <- ggplot(data=injuries, aes(x=reorder(evtype, injuries), y=injuries, color=evtype)) + 
  geom_bar(stat="identity",fill="white") + 
  xlab("Event Type") + 
  ylab("Total number injuries") +  
  ggtitle("10 Highest injuries  Events") +
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"))

injuriesPlot + coord_flip()
```



```{r,echo = TRUE} 

fatalitiesPlot <- ggplot(data=fatalities, aes(x=reorder(evtype, fatalities), y=fatalities, color=evtype)) + 
  geom_bar(stat="identity",fill="white") + 
  xlab("Event Type") + 
  ylab("Total number fatalities ") +  
  ggtitle("10 Highest fatalities  Events") +
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"))

fatalitiesPlot + coord_flip()
```


```{r,echo = TRUE} 
propdamagePlot <- ggplot(data=propdamage, aes(x=reorder(evtype, propdmgvalue), y=propdmgvalue, color=evtype)) + 
  geom_bar(stat="identity",fill="white") + 
  xlab("Event Type") + 
  ylab("Total Property in dollars") +  
  ggtitle("10 Highest Property Damages Events") +
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"))

propdamagePlot + coord_flip()
```


```{r,echo = TRUE} 
cropdamagePlot <- ggplot(data=cropdamage, aes(x=reorder(evtype, cropdmgvalue), y=cropdmgvalue, color=evtype)) + 
  geom_bar(stat="identity",fill="white") + 
  xlab("Event Type") + 
  ylab("Total Crop in dollars") +  
  ggtitle("10 Highest Crop Damages Events") +
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"))

cropdamagePlot + coord_flip()
```




