---
title: "Reproducible Research - Course Project 2"
author: "echf"
date: "11/8/2021"
output: html_document
---




```{r,echo = TRUE}
options(scipen = 1)  # Turn off scientific notations for numbers
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)
library(dplyr)
```

```{r,echo = TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "StormData.csv.bz2")
library(R.utils)
bunzip2("StormData.csv.bz2", "StormData.csv", overwrite=T, remove=F)
df_storm <- read.csv("StormData.csv")
names(df_storm)
str(df_storm)
```




#### Economic variables:
  
#####  PROPDMG: approx. property damags
#####  PROPDMGEXP: the units for property damage value
#####  CROPDMG: approx. crop damages
#####  CROPDMGEXP: the units for crop damage value
 
  
####   Events - target variable:
#####    EVTYPE: weather event (Tornados, Wind, Snow, Flood, etc..)
#####    Extract variables of interest from original data set:

####  Health variables:
  * INJURIES: approx. number of injuries
  * FATALITIES: approx. number of deaths
  * propdamage
  * cropdamage

```{r,echo = TRUE}
head(df_storm,10)
```


#### Extracting variables of interest for analysis of weather impact on health and economy
#### From a list of variables in storm.data, these are columns of interest:
  


```{r,echo = TRUE}
df_storm2 <- df_storm %>% 
  select(c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")) %>%
  rename_all(tolower)
str(df_storm2)
```


<!-- Property damage estimates were entered as actual dollar amounts (the variable PROPDMG). But they were rounded to three significant digits, followed by an alphabetical character signifying the magnitude of the number, i.e., 1.55B for $1,550,000,000. Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions. So I created a new variable PROPDMGEXP2 and assigned conditionally "K" = 1000, "M" = 1000000, "B" = 1000000000, in other cases 1. These variables are multiplied in the next step. -->

```{r,echo = TRUE}
sort(table(df_storm2$propdmgexp), decreasing = TRUE)[1:10]

sort(table(df_storm2$cropdmgexp), decreasing = TRUE)[1:10]

```
<!-- There is some mess in units, so we transform those variables in one unit (dollar) variable by the following rule: -->
<!-- * K or k: thousand dollars (10^3) -->
<!-- * M or m: million dollars (10^6) -->
<!-- * B or b: billion dollars (10^9) -->
<!-- * the rest would be consider as dollars -->

<!-- Transform EVTYPE to all caps -->
<!-- This will improve consistency across categories -->

```{r,echo = TRUE}
df_storm2$evtype <- as.character(df_storm2$evtype) #required to transform
df_storm2$evtype <- toupper(df_storm2$evtype)
```

<!-- Calculate property damage value -->
<!-- Need to recode the exponent variable so we can multiply it by the propdmg value. -->

```{r,echo = TRUE}  
df_storm2$propdmgexp <- toupper(df_storm2$propdmgexp)
df_storm2 <- mutate(df_storm2, propdmgexp2 = recode(propdmgexp,
                                             "1" = 10,
                                             "2" = 100,
                                             "3" = 1000,
                                             "4" = 10000,
                                             "5" = 100000,
                                             "6" = 1000000,
                                             "7" = 10000000,
                                             "8" = 100000000,
                                             "H" = 100,
                                             "K" = 1000,
                                             "M" = 1000000,
                                             "B" = 1000000000,
                                             .default = 1)) 
df_storm2 <- mutate(df_storm2, propdmgvalue = propdmg*propdmgexp2)
```



<!-- Calculate crop damage value -->
<!-- Need to recode the exponent variable so we can multiply it by the cropdmg value. -->
  
```{r,echo = TRUE}  
df_storm2$cropdmgexp <- toupper(df_storm2$cropdmgexp)
df_storm2 <- mutate(df_storm2, cropdmgexp2 = recode(cropdmgexp,  
                                                  "1" = 10,
                                                  "2" = 100,
                                                  "3" = 1000,
                                                  "4" = 10000,
                                                  "5" = 100000,
                                                  "6" = 1000000,
                                                  "7" = 10000000,
                                                  "8" = 100000000,
                                                  "H" = 100,
                                                  "K" = 1000,
                                                  "M" = 1000000,
                                                  "B" = 1000000000,
                                                 .default = 1))
df_storm2 <- mutate(df_storm2, cropdmgvalue = cropdmg*cropdmgexp2)
```

<!-- Subset data for analyses -->
<!-- Injury data -->
  
```{r,echo = TRUE}  
injuries <- aggregate(injuries~evtype, data = df_storm2, FUN = sum) 
injuries <- injuries[order(injuries$injuries, decreasing = TRUE), ] 
injuries <- injuries[1:5, ] 
```

<!-- Fatality data -->
```{r,echo = TRUE}  
fatalities <- aggregate(fatalities~evtype, data = df_storm2, FUN = sum) #aggregate fatalities by event type
fatalities <- fatalities[order(fatalities$fatalities, decreasing = TRUE), ] #order number of fatalities in decending order
fatalities <- fatalities[1:5, ] #take the top 5
```

<!-- Property damage data -->
```{r,echo = TRUE}
propdamage <- aggregate(propdmgvalue~evtype, data = df_storm2, FUN = sum)
propdamage <- propdamage[order(propdamage$propdmgvalue, decreasing = TRUE), ]
propdamage <- propdamage[1:5, ]
```

<!-- Crop damage data -->
```{r,echo = TRUE}  
cropdamage <- aggregate(cropdmgvalue~evtype, data = df_storm2, FUN = sum)
cropdamage <- cropdamage[order(cropdamage$cropdmgvalue, decreasing = TRUE), ]
cropdamage <- cropdamage[1:5, ]
```







